Abstraction and Invariance for Algebraically Indexed Types

### 2012-07-10 paper/geometry-twocolumn.tex
Patty's changes to shorten the text

\begin{example}\label{ex:uninhabited-type}
  So far we have emphasised the derivation of properties, or ``free
  theorems'', of programs from their types. But using more refined
  relational interpretations of types we can also show that certain
  types are uninhabited. For example, the type $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
  quantifier-free inhabitants\footnote{We define quantifier-free
    programs in the statement of
    \thmref{thm:abstraction}.}. Intuitively, this is because there is
  no way to remove the extra $t$ in $\tyPrim{vec}{t+t}$ using the
  vector operations. We formalise this indefinability result in
  \autoref{sec:types-indexed-abelian-groups-indef} using a more
  sophisticated relational interpretation of free translation
  variables than the one we sketched above.
\end{example}

>>>>

\begin{example}\label{ex:uninhabited-type}
  So far we have emphasised the derivation of properties, or ``free
  theorems'', of programs from their types. But using more refined
  relational interpretations of types we can also show that certain
  types are uninhabited. For example, the type $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$ has no
  quantifier-free inhabitants\footnote{We define quantifier-free
    programs in the statement of
    \thmref{thm:abstraction}.}. Intuitively, this is because we cannot
  remove the extra $t$ in $\tyPrim{vec}{t+t}$ using the
  vector operations. We formalise this indefinability result in
  \autoref{sec:types-indexed-abelian-groups-indef} using a more
  sophisticated relational interpretation of free translation
  variables than the one sketched above.
\end{example}

### 2012-07-10 paper/section-discussion.tex
Patty's changes to shorten the text

We have presented a general framework for algebraically indexed types
and instantiated it to yield novel type systems for geometry, logical
information flow and distance-indexed types. Our framework further
demonstrates the power of relational reasoning about typed
programs. From \thmref{thm:abstraction}, we have derived interesting
free theorems, type isomorphisms and non-definability results. We
conclude with some observations and suggestions for further work.

\paragraph{Further Applications and Extensions} We have covered
several applications of algebraically indexed types in this paper, but
there are undoubtedly many more. Geometry for dimensions greater than
two is an obvious candidate, as well as systems that are invariant
under different geometric groups (e.g.,~the Poincar\'{e} group for
relativity). Mathematical Physics is particularly rich in theories
that have some notion of invariance, and it will be exciting to pin
down the precise connections between these and type systems for which
analogues of Reynolds' abstraction theorem hold. Geometric theorem
proving is another possible application area. Harrison
\cite{harrison09without} comments on the pervasive useful of
invariance properties in this area. Programs in our framework
automatically satisfy invariance properties, removing the need for
ad-hoc proofs of these facts.

@@

Type and effect analyses use types indexed by effect annotations that
have algebraic structure (e.g.~sets of read/write effect labels, with
the structure of an idempotent monoid). Benton et al.~have used
relational interpretations to prove effect-dependent equivalences
\cite{benton06reading}. We expect that an extension of our framework
with type-indexed types should be able to express their effect-indexed
monads and prove their equivalences.

>>>>

We presented a general framework for algebraically indexed types
and instantiated it to yield novel type systems for geometry, taint
tracking, logical information flow and distance-indexed types. Our
framework further demonstrates the power of relational reasoning about
typed programs. From \thmref{thm:abstraction}, we derived
interesting free theorems, type isomorphisms and non-definability
results. We conclude with some observations and suggestions for
further work.

\paragraph{Further Applications and Extensions} We have covered
several applications of algebraically indexed types in this paper, but
there are undoubtedly many more. Geometry for dimensions greater than
two is an obvious candidate, as are systems that are invariant
under different geometric groups (e.g.,~the Poincar\'{e} group for
relativity). Mathematical Physics is particularly rich in theories
that have some notion of invariance, and it will be exciting to pin
down the precise connections between these and type systems for which
%analogues of Reynolds' 
an abstraction theorem holds. Geometric theorem
proving is another %possible 
application. Harrison
\cite{harrison09without} comments on the pervasiveness of % usefulness of
invariance properties in this area. Programs in our framework
automatically satisfy invariance properties, removing the need for
{\em ad hoc} proofs. % of these facts.

@@

Type and effect analyses use types indexed by effect annotations with
algebraic structure (e.g.,~sets of read/write effect labels with an
idempotent monoid structure). Benton {\em et al.}~ have used relational
interpretations to prove effect-dependent equivalences
\cite{benton06reading}. An extension of our framework with
type-indexed types should be able to express their effect-indexed monads
and prove their equivalences.

### 2012-07-10 paper/section-general-framework.tex
Inhabitation of AG
Space saving
Patty's changes to shorten the text

  \bigskip

@@

Given a set $\IndexAxiomSet$ of axioms, we generate the equality
judgment between index expressions $\Delta \vdash e \equiv e' : s$ by
a set of rules. The following rule allows us to use substitution
instances of axioms.
\begin{displaymath}
  \inferrule*
  {(\Delta' \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
    \Delta \vdash \sigma \Rightarrow \Delta'}
  {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}
\end{displaymath}
We also assume the standard congruence, reflexivity, symmetry and
transitivity rules for the equality judgment.

@@

\paragraph{Non-compositional Relational Environments}
We now construct relational environments that can be used to prove
non-definability results. To this end, consider the typing
$t\mathord:\SynTransl{2}; \Gamma_{\Geom} \vdash M : \tyPrim{vec}{t +
  t} \to \tyPrim{vec}{t}$ involving vectors indexed by translations.
In \exref{ex:uninhabited-type} we claimed that this type is
uninhabited. Intuitively, it is clear why: given the operations on
vectors we presented in \autoref{sec:affine-vector-ops}, there is no
way of cancelling the additional $t$ in the argument type to get the
result type. More generally, our operations on vectors only allow us
to combine indices according to the group operations. Thus, if the
index of the result type of a function does not lie within the free
abelian group generated by the indices of the input types, then there
is no way to write a program with this %given
type. We now develop this idea to generate a family of sets of
relational environments. The key idea is to assign different
relational interpretations to types with indices that lie outside some
finitely generated collection of index expressions.

>>>>

  \smallskip

@@

Given a set $\IndexAxiomSet$ of axioms, we generate the equality
judgment between index expressions $\Delta \vdash e \equiv e' : s$ by
a set of rules. The following rule lets us use substitution
instances of axioms:
\begin{displaymath}
  \inferrule*
  {(\Delta' \vdash e \stackrel{ax}\equiv e' : s) \in \IndexAxiomSet \\
    \Delta \vdash \sigma \Rightarrow \Delta'}
  {\Delta \vdash \sigma^*e \equiv \sigma^*e' : s}
\end{displaymath}
We also assume the standard congruence, reflexivity, symmetry and
transitivity rules for the equality judgment.

@@

\paragraph{Non-compositional Relational Environments}
We now construct relational environments that can be used to prove
non-definability results. To this end, consider the typing
$t\mathord:\SynTransl{2}; \Gamma_{\Geom} \vdash M : \tyPrim{vec}{t +
  t} \to \tyPrim{vec}{t}$ involving vectors indexed by translations.
In \exref{ex:uninhabited-type} we claimed that this type is
uninhabited. Intuitively, it is clear why: given the operations on
vectors in \autoref{sec:affine-vector-ops}, there is no way of
cancelling the additional $t$ in the argument type to get the result
type. More generally, the vector operations only allow us to combine
indices according to the group operations, so if the index of the
result type of a function lies outside the free abelian group
generated by the indices of the input types, then we cannot write a
program of this type. 
We now develop this idea to generate a
family of sets of relational environments. The key idea is to assign
different relational interpretations to types with indices that lie
outside some finitely generated collection of index expressions.

### 2012-07-10 paper/section-info-flow.tex
Patty's changes to shorten the text

Using algebraic indexing, we refine Tse and Zdancewic's translation by
replacing each type variable $\alpha_l$ with a primitive type
$\tyPrim{T}{e}$ of representations of the truth of a logical
proposition $e$. The relationships between security levels are now
replaced by logical entailment, so we only have functions of type
$\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
$\psi$. Inspired by the PER interpretation of information flow, we
instantiate our general relational framework to interpret
$\tyPrim{T}{\phi}$ as the identity relation when $\phi$ is true, and
the empty relation when $\phi$ is false. By doing so, we recover the
information flow properties of Abadi et al. and Tse and Zdancewic.

>>>>

Using algebraic indexing, we refine Tse and Zdancewic's translation by
replacing each type variable $\alpha_l$ with a primitive type
$\tyPrim{T}{e}$ of representations of the truth of a logical
proposition $e$. The relationships between security levels are now
replaced by logical entailment, so we only have functions of type
$\tyPrim{T}{\phi} \to \tyPrim{T}{\psi}$ when $\phi$ entails
$\psi$. %Inspired by the PER interpretation of information flow, 
We instantiate our general relational framework to interpret
$\tyPrim{T}{\phi}$ as the identity relation if $\phi$ is true and the
empty relation if $\phi$ is false. This recovers the information flow
properties of Abadi {\em et al.} and Tse and Zdancewic.

### 2012-07-10 paper/section-instantiations.tex
Patty's changes to shorten the text

\subsection{Indefinability}
\label{sec:types-indexed-abelian-groups-indef}
\newcommand{\Grp}{\mathit{Grp}}
For proving non-definability results, we
consider the general case of a sort $\mathsf{G}$ with the equational theory of abelian groups.
We assume a group-indexed primitive type $\tyPrimNm{G}$ and
set of primitive operations typed by $\Gamma_\Grp$ that reflect the group structure at the value level:
\begin{displaymath}
  \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
    1 & \tyPrim{G}{1} \\
    (\cdot) & \forall a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to \tyPrim{G}{a\cdot b} \\
    \ ^{-1} & \forall a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}
  \end{array}
\end{displaymath}
One instance is the sort $\SynTransl{2}$ with $0$, $(+)$ and $\mathrm{negate}$ operations.

For index-erasure interpretation, we assume a (not necessarily abelian) group $(G, 1, \cdot,
\ ^{-1})$. Let $Z(G)$ be the centre of the $G$: i.e.,~the set of
elements that commute with every other element in $G$. We set
$\tyPrimSem{G} = G$. The primitive operations in $\Gamma_\Grp$ are
interpreted just as the unit, multiplication and inverse operations
of $G$. For each index context $\Delta$, the set
$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
\to Z(G) \}$, where
\begin{displaymath}
  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
    \begin{array}{ll}
      \{ \} & \textrm{if}\ e\not\in H \\
      \{ (x, h(e)\oplus x) \sepbar x \in G \} & \textrm{if}\ e\in H
    \end{array}
  \right.
\end{displaymath}

@@

We write $i^n$ for $n$-fold iteration of the group operation.
\begin{theorem}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
  there exists a program $M$ such that
  \begin{equation}
    \label{eq:fo-group-type}
    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
  \end{equation}
  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, if and only if there is a solution $\vec{x}$
  in integers to the equation $A \vec x = b$.
\end{theorem}
The uninhabited type  $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$
from \exref{ex:uninhabited-type} is an instance of this theorem.
For types indexed by the scaling group $\SynGL{1}$ the situation is more subtle, as
the index-erasure interpretation of $\tyPrim{real}{s}$ is group $\GL{1}$ together with zero.
We can show, nonetheless, that $\forall s \mathord:
  \SynGL{1}.\ \tyPrim{real}{s \cdot s} \to \tyPrim{real}{s}\cong 1$, reflecting 
the fact that any term of this type must be $\lambda x.0$.

>>>>

\subsection{Indefinability}
\label{sec:types-indexed-abelian-groups-indef}
\newcommand{\Grp}{\mathit{Grp}} To prove non-definability results,
consider %the general case of 
a sort $\mathsf{G}$ with the equational
theory of abelian groups.  We assume a group-indexed primitive type
$\tyPrimNm{G}$ and set of primitive operations typed by $\Gamma_\Grp$
that reflect the group structure at the value level, namely $1 :
\tyPrim{G}{1}$, $ (\cdot) : \forall
a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
\tyPrim{G}{a\cdot b}$, and $\ ^{-1} : \forall
a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{a^{-1}}$.
%\begin{displaymath} \begin{array}{r@{\hspace{0.5em}:\hspace{0.5em}}l}
%  1 & \tyPrim{G}{1} \\ (\cdot) & \forall
%  a,b\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to \tyPrim{G}{b} \to
%  \ty%Prim{G}{a\cdot b} \\ \ ^{-1} & \forall
%  a\mathord:\mathsf{G}.\ \tyPrim{G}{a} \to
%  \tyPrim{G}{a^{-1}} \end{array} \end{displaymath}
One instance is the sort $\SynTransl{2}$ with $0$, $(+)$ and
$\mathrm{negate}$ operations.

For index-erasure interpretation, we assume a (not necessarily
abelian) group $(G, 1, \cdot, \ ^{-1})$. Let $Z(G)$ be the centre of
the $G$ %: i.e.,~the set of elements that commute with every other
%element in $G$. We set 
and $\tyPrimSem{G} = G$. The primitive operations
in $\Gamma_\Grp$ are interpreted just as the unit, multiplication and
inverse operations of $G$. For each index context $\Delta$, the set
$\relEnv{E}_\Grp(\Delta)$ of relational environments is defined to be
$\{ \rho_{(H,h)} \sepbar H\textrm{ is a subgroup of }(\Delta
\Rightarrow \mathsf{G})/\equiv, h\textrm{ is a group homomorphism }H
\to Z(G) \}$, where
\begin{displaymath}
  \rho_{(H,h)}\ \tyPrimNm{G}\ (e) = \left\{
    \begin{array}{ll}
      \{ \} & \textrm{if}\ e\not\in H \\
      \{ (x, h(e)\oplus x) \sepbar x \in G \} & \textrm{if}\ e\in H
    \end{array}
  \right.
\end{displaymath}

@@

Writing $i^n$ for $n$-fold iteration of the group operation, we have:
\begin{theorem}
  Let $m$ and $n$ be natural numbers. For all $n$-by-$m$-matrices of
  natural numbers $A$ and $m$-by-$1$ matricies of integers $b$,
  there exists a program $M$ such that
  \begin{equation}
    \label{eq:fo-group-type}
    \Gamma_\Grp \vdash M : \forall i_1,\dots,i_m\mathord:\mathsf{G}.\ \tyPrim{G}{e_1} \to \dots \to \tyPrim{G}{e_n} \to \tyPrim{G}{e}
  \end{equation}
  where $e_j = i_1^{A_{j1}}\cdots i_m^{A_{jm}}$ and $e = i_1^{b_1}\cdots i_m^{b_m}$, if and only if there is a solution $\vec{x}$
  in integers to the equation $A \vec x = b$.
\end{theorem}
The uninhabited type  $\forall t \mathord:
  \SynTransl{2}.\ \tyPrim{vec}{t + t} \to \tyPrim{vec}{t}$
from \exref{ex:uninhabited-type} is an instance of this theorem.
For types indexed by the scaling group $\SynGL{1}$ the situation is more subtle, as
the index-erasure interpretation of $\tyPrim{real}{s}$ is group $\GL{1}$ together with zero.
We can show, nonetheless, that $\forall s \mathord:
  \SynGL{1}.\ \tyPrim{real}{s \cdot s} \to \tyPrim{real}{s}\cong 1$, reflecting 
the fact that any term of this type must be $\lambda x.0$.

### 2012-07-10 paper/section-introduction.tex
Lots of little typesetting fixes and typos
Patty's changes to shorten the text

As it happens, the above examples can all be expressed using unary
functions rather than binary relations: we can write functions that
scale, or translate, one value to another according to their type,
even for higher-order function types. Moreover, the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\beta) =
\rho(\alpha)\cdot\rho(\beta)$ where product of units-of-measure is
interpreted by product of scale factors. For more sophisticated
applications, however, we need the fully generality of binary
relations. As is the case with type polymorphism, it
provides the appropriate meaning for nested quantifiers in types such
as
$(\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
More interestingly, particular applications sometimes require
relations that are not built compositionally---we can, for instance,
use such relations to prove that there are no non-trivial programs of
type
$\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
Finally, there are interesting applications in which even the basic
relations are not functional. In a type system in which the index in
$\tyPrim{real}\alpha$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(\alpha)$ for a positive real number
$\rho(\alpha)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as a type:
\begin{displaymath}
  \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
\end{displaymath}

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a type
system for tracking taint marking (or provenance) of strings, a
refined type system for information flow based on logic, and a simple
type system for basic uniform continuity analysis.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' Abstraction Theorem. We give precise conditions
on sets of relation environments under which our abstraction theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and Abstraction Theorem in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}, available from
\\\url{github.com/bobatkey/algebraically-indexed-types/tree/master/coq}

>>>>

As it happens, the above examples can all be expressed using unary
functions rather than binary relations: we can write functions that
scale, or translate, one value to another according to their type,
even for higher-order function types. Moreover, the interpretation of
indices can be built up compositionally, by recursion over the
syntactic structure. For example, we have $\rho(\alpha\cdot\beta) =
\rho(\alpha)\cdot\rho(\beta)$ where product of units-of-measure is
interpreted by product of scale factors. For more sophisticated
applications, however, we need the fully generality of binary
relations. As is the case with type polymorphism, it
provides the appropriate meaning for nested quantifiers in types such
as
$(\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}\alpha\to\tyPrim{real}\alpha)\to\tyBool$.
More interestingly, particular applications sometimes require
relations that are not built compositionally---we can, for instance,
use such relations to prove that there are no non-trivial programs of
type
$\forall\alpha\mathord:\SynGL{1}.\tyPrim{real}{\alpha\cdot\alpha}\to\tyPrim{real}\alpha$.
Finally, there are interesting applications in which even the basic
relations are not functional. In a type system in which the index in
$\tyPrim{real}\alpha$ is interpreted not as a unit of measure, but as
a measure of \emph{closeness}, two values $x$ and $y$ of this type are
related if $|x-y| < \rho(\alpha)$ for a positive real number
$\rho(\alpha)$.  Rather beautifully, the standard notion of uniform
continuity can then be expressed as %a type:
%\begin{displaymath}
 $ \forall \epsilon \mathord: \mathsf{R}^{>0}.\ \exists \delta\mathord: \mathsf{R}^{>0}.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}

@@

This paper makes the following specific contributions:
\begin{itemize}
\item 
We present a collection of compelling examples of algebraically
indexed types, including a novel type system for geometry, a type
system for tracking taint marking (or provenance) of strings, a
refined type system for information flow based on logic, and a simple
type system for basic uniform continuity analysis.
\item 
We formulate a type system that can either be used as a programming
language in its own right, or as the target of type-based
analyses. The type system consists of the usual type constructors
together with a collection of indexed primitive types, universal and
existential quantification over the indices, and a multi-sorted
equational theory for indices.
\item
We describe a relational semantics for the type system and prove an
analogue of Reynolds' Abstraction Theorem. We give precise conditions
on sets of relation environments under which our abstraction theorem
holds.
\item
For each of our main examples we deduce free theorems that are
consequences of our abstraction theorem, prove non-definability
results, and derive interesting type isomorphisms.
\end{itemize}
We have fully formalised our framework and Abstraction Theorem in Coq,
using strongly-typed term representations
throughout~\cite{TypedSyntax}. The formalisation is available from
\\{\small \url{https://github.com/bobatkey/algebraically-indexed-types}}

### 2012-07-10 paper/section-metric.tex
Patty's changes to shorten the text

\paragraph{Uniform continuity}
  %By making use of 
Using %the
existential types, % in our framework,
we can state the standard $\epsilon$-$\delta$ definition of
\emph{uniform} continuity as a type:
\begin{displaymath}
  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}
\end{displaymath}
Given a program $M$ with this type, we obtain from
\thmref{thm:abstraction} a free theorem, which is exactly the
definition of uniform continuity:
\begin{displaymath}
  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. |x - x'| < \delta \Rightarrow |\tmSem{M}\ x - \tmSem{M}\ x'| < \epsilon
\end{displaymath}
This definition differs from the $\epsilon$-$\delta$ definition of
(regular) continuity by the order of quantification:
  %For continuity, 
there, $\forall x$ comes before $\exists \delta$, so the distance
$\delta$ may depend on %the value of
$x$. We suspect that to express standard continuity as a type would
require some form of type dependency. Chaudhuri, Gulwani and
Lublinerman \cite{chaudhuri10continuity} have given a program logic
based approach to verifying the continuity of programs.

>>>>

\paragraph{Uniform continuity}
  %By making use of 
Using %the
existential types, % in our framework,
%we can state 
the standard $\epsilon$-$\delta$ definition of
\emph{uniform} continuity can be expressed as %a type:
%\begin{displaymath}
$  \forall \epsilon \mathord: \metricSort.\ \exists \delta\mathord: \metricSort.\\ \tyPrim{real}{\delta} \to \tyPrim{real}{\epsilon}$.
%\end{displaymath}
For any program $M$ of this type, %we obtain from
\thmref{thm:abstraction} gives a free theorem that is exactly 
%the definition of 
uniform continuity:
\begin{displaymath}
  \forall \epsilon \mathord> 0. \exists \delta \mathord> 0. \forall x,x'. |x - x'| < \delta \Rightarrow |\tmSem{M}\ x - \tmSem{M}\ x'| < \epsilon
\end{displaymath}
This definition differs from the $\epsilon$-$\delta$ definition of
(regular) continuity by the order of quantification:
  %For continuity, 
there, $\forall x$ comes before $\exists \delta$, so the distance
$\delta$ may depend on %the value of
$x$. We suspect that to express standard continuity as a type would
require some form of type dependency. Chaudhuri, Gulwani and
Lublinerman \cite{chaudhuri10continuity} have given a program logic
based approach to verifying the continuity of programs.
